---
title: "Introduction to dplyr"
output:
  html_document:
    df_print: paged
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
library(dplyr)
library(ggplot2)
set.seed(1014)
```

When working with data you must:

* Figure out what you want to do.

* Describe those tasks in the form of a computer program.

* Execute the program.

The dplyr package makes these steps fast and easy:

* By constraining your options, it helps you think about your data manipulation 
  challenges.

* It provides simple "verbs", functions that correspond to the most common data 
  manipulation tasks, to help you translate your thoughts into code.

* It uses efficient backends, so you spend less time waiting for the computer.

This document introduces you to dplyr's basic set of tools, and shows you how to apply them to data frames. dplyr also supports databases via the dbplyr package, once you've installed, read `vignette("dbplyr")` to learn more.

## Data: nycflights13

To explore the basic data manipulation verbs of dplyr, we'll use `nycflights13::flights`. This dataset contains all `r nrow(nycflights13::flights)` flights that departed from New York City in 2013. The data comes from the US [Bureau of Transportation Statistics](http://www.transtats.bts.gov/DatabaseInfo.asp?DB_ID=120&Link=0), and is documented in `?nycflights13`

```{r}
library(nycflights13)
dim(flights)
flights
```

Note that `nycflights13::flights` is a tibble, a modern reimagining of the data frame. It's particularly useful for large datasets because it only prints the first few rows. You can learn more about tibbles at <http://tibble.tidyverse.org>; in particular you can convert data frames to tibbles with `as_tibble()`.

## Single table verbs

Dplyr aims to provide a function for each basic verb of data manipulation:

* `filter()` to select cases based on their values.
* `arrange()` to reorder the cases.
* `select()` and `rename()` to select variables based on their names.
* `mutate()` and `transmute()` to add new variables that are functions of existing variables.
* `summarise()` to condense multiple values to a single value.
* `sample_n()` and `sample_frac()` to take random samples.

### Filter rows with `filter()`

`filter()` allows you to select a subset of rows in a data frame. Like all single verbs, the first argument is the tibble (or data frame). The second and subsequent arguments refer to variables within that data frame, selecting rows where the expression is `TRUE`.

For example, we can select all flights on January 1st with:

```{r}
filter(flights, month == 1, day == 1)
```

Do the equivalent in base R code:

```{r, eval = FALSE}
flights[flights$month == 1 & flights$day == 1, ]
```

### Arrange rows with `arrange()`

`arrange()` works similarly to `filter()` except that instead of filtering or selecting rows, it reorders them. It takes a data frame, and a set of column names (or more complicated expressions) to order by. If you provide more than one column name, each additional column will be used to break ties in the values of preceding columns.

Arrange flights by date (so year, month and day):

```{r}
arrange(flights, year, month, day)
```

Use `desc()` to order a column in descending order of arr_delay:

```{r}
arrange(flights, desc(arr_delay))
```

### Select columns with `select()`

Often you work with large datasets with many columns but only a few are actually of interest to you. `select()` allows you to rapidly zoom in on a useful subset using operations that usually only work on numeric variable positions:

```{r}
# Select columns year, month and day by name
select(flights, year, month, day)
# Select all columns between year and day (inclusive)
select(flights, year:day)
# Select all columns except those from year to day (inclusive)
select(flights, -(year:day))
```

There are a number of helper functions you can use within `select()`, like `starts_with()`, `ends_with()`, `matches()` and `contains()`. These let you quickly match larger blocks of variables that meet some criterion. See `?select` for more details.

You can rename variables with `select()` by using named arguments.

Rename the tailnum to tail_num using `select()`

```{r}
select(flights, tail_num = tailnum)
```

But because `select()` drops all the variables not explicitly mentioned, it's not that useful. Instead, use `rename()` to rename tailnum:

```{r}
rename(flights, tail_num = tailnum)
```

### Add new columns with `mutate()`

Besides selecting sets of existing columns, it's often useful to add new columns that are functions of existing columns.  This is the job of `mutate()`.

Use mutate to create two new variables:

* **gain** the arrival delay less the departure delay
* **speed** the distance divided by the air time

```{r}
mutate(flights,
  gain = arr_delay - dep_delay,
  speed = distance / air_time * 60
)
```


### Summarise values with `summarise()`

The last verb is `summarise()`. It collapses a data frame to a single row.

Summarise the flights data frame to just the mean departure delay

```{r}
summarise(flights,
  delay = mean(dep_delay, na.rm = TRUE)
)
```

It's not that useful until we learn the `group_by()` verb below.

You use `summarise()` with __aggregate functions__, which take a vector of values and return a single number. There are many useful examples of such functions in base R like `min()`, `max()`, `mean()`, `sum()`, `sd()`, `median()`, and `IQR()`. dplyr provides a handful of others:

* `n()`: the number of observations in the current group

* `n_distinct(x)`:the number of unique values in `x`.

* `first(x)`, `last(x)` and `nth(x, n)` - these work
  similarly to `x[1]`, `x[length(x)]`, and `x[n]` but give you more control
  over the result if the value is missing.

For example, we could use these to find the number of planes and the number of flights that go to each possible destination.

Create a new dataframe `destinations` by grouping `flights` by `dest`

Then summarise `destinations` to get `planes`, the number of unique planes at each destination and `flights` the number of flights to each destination.

```{r}
destinations <- group_by(flights, dest)
summarise(destinations,
  planes = n_distinct(tailnum),
  flights = n()
)
```
